### 2022-05-10 paper/localpsf.tex
minor edits
wrote about recycling krylov information

Approximation \eqref{eq:translate1} says that $\phi_x$ can be approximated by $\phi_{x'}$ when $x$ and $x'$ are close. Operators satisfying \eqref{eq:translate1} can be well-approximated via low-rank interpolative factorization/CUR approximation \cite{CUR,NYSTROM}. However, the required rank in the low rank approximation can be large, which makes algorithms based on \eqref{eq:translate1} expensive.

@@

\section{Fast ellipsoid intersection test}
\label{sec:fast_ellipsoid_intersection_test}
The procedure for choosing sample points relies on quickly determining whether two ellipsoids intersect. Let $E_1$ and $E_2$ be the ellipsoids defined as
\begin{align*}
	E_1 :=& \{x : (x - \spatialmean_1)^T \spatialcov_1^{-1} (x - \spatialmean_1) \le \tau^2\} \\
	E_2 :=& \{x : (x - \spatialmean_2)^T \spatialcov_2^{-1} (x - \spatialmean_2) \le \tau^2\}, \\
\end{align*}
where $\spatialmean_1, \spatialmean_2 \in \mathbb{R}^\gdim$, and $\spatialcov_1, \spatialcov_2 \in \mathbb{R}^{\gdim \times \gdim}$ are positive definite. Let $K$ be the following one dimensional convex function:
\begin{equation*}
	K(s) := 1 - \frac{1}{\tau^2} (\spatialmean_1 - \spatialmean_2)^T \left(\frac{1}{1-s}\spatialcov_1 + \frac{1}{s}\spatialcov_2\right)^{-1}(\spatialmean_1 - \spatialmean_2)	
\end{equation*}
In \cite{GilitschenskiHanebeck12} it is shown that $E_1 \cap E_2 = \{\}$ if and only if $K(s) < 0$ for some $s\in (0,1)$. We check whether $E_1$ and $E_2$ intersect by minimizing $K(s)$ on $(0,1)$. If $K(s^*) <0$ at the minimizer $s^*$, then $E_1 \cap E_2 = \{\}$. Otherwise $E_1 \cap E_2 \neq \{\}$.

@@

>>>>

Approximation \eqref{eq:translate1} says that $\phi_x$ can be approximated by $\phi_{x'}$ when $x$ and $x'$ are close. Operators satisfying \eqref{eq:translate1} can be well-approximated via low-rank CUR approximation \cite{CUR,NYSTROM}. However, the required rank in the low rank approximation can be large, which makes algorithms based on \eqref{eq:translate1} expensive.

@@

\section{Ellipsoid intersection test}
\label{sec:fast_ellipsoid_intersection_test}
The procedure for choosing sample points relies on quickly determining whether two ellipsoids intersect. Let $E_1$ and $E_2$ be the ellipsoids defined as
\begin{align*}
	E_1 :=& \{x : (x - \spatialmean_1)^T \spatialcov_1^{-1} (x - \spatialmean_1) \le \tau^2\} \\
	E_2 :=& \{x : (x - \spatialmean_2)^T \spatialcov_2^{-1} (x - \spatialmean_2) \le \tau^2\}, \\
\end{align*}
where $\spatialmean_1, \spatialmean_2 \in \mathbb{R}^\gdim$, and $\spatialcov_1, \spatialcov_2 \in \mathbb{R}^{\gdim \times \gdim}$ are positive definite. Let $K$ be the following one dimensional convex function:
\begin{equation*}
	K(s) := 1 - \frac{1}{\tau^2} (\spatialmean_1 - \spatialmean_2)^T \left(\frac{1}{1-s}\spatialcov_1 + \frac{1}{s}\spatialcov_2\right)^{-1}(\spatialmean_1 - \spatialmean_2)	
\end{equation*}
In \cite{GilitschenskiHanebeck12} it is shown that $E_1 \cap E_2 = \{\}$ if and only if $K(s) < 0$ for some $s\in (0,1)$. We check whether $E_1$ and $E_2$ intersect by minimizing $K(s)$ on $(0,1)$. If $K(s^*) <0$ at the minimizer $s^*$, then $E_1 \cap E_2 = \{\}$. Otherwise $E_1 \cap E_2 \neq \{\}$.

@@

\section{Recycling Krylov information via low rank updates}
\label{sec:recycle_krylov}

In applications, often one wants to use the H-matrix approximation $\mathbf{A}_H$ to build preconditioners for solving a sequence of slowly varying linear systems of the form
\begin{equation}
\label{eq:Ax_equals_b}
\mathbf{B} \mathbf{u} = \mathbf{f},
\end{equation}
where each of these linear systems is solved using an iterative method. If $\mathbf{B}=\mathbf{A}$, then $\mathbf{A}_H^{-1}$ may be used to precondition this linear system. If $\mathbf{B}$ is built from a combination of $\mathbf{A}$ and other operators, then $\mathbf{A}_H$ may be combined with H-matrix approximations of the other operators to construct a preconditioner for $\mathbf{B}$. For example, in a Newton-Krylov method for solving an optimization problem, one uses a Krylov method such as conjugate gradient to solve a sequence of systems of the form \eqref{eq:Ax_equals_b}, where $\mathbf{B}$ is the Hessian of the objective function in the optimization problem, and $\mathbf{f}$ is the negative gradient of the objective function. Both the Hessian and the gradient are evaluated at the current Newton iterate, so $\mathbf{B}$ and $\mathbf{f}$ change as the Newton iterations proceed. The Hessian is typically the sum of a data-misfit term which is available matrix-free, and a regularization term which is available directly as a sparse matrix. The data misfit term in the Hessian is approximated with our method to form $\mathbf{A}_H$, the regularization term is approximated using standard H-matrix methods for sparse matrices, and then these two terms are added together and inverted using H-matrix methods to form a preconditioner for the overall Hessian.

Let 
\begin{equation*}
\mathbf{P} \approx \mathbf{B}^{-1}
\end{equation*}
denote the H-matrix preconditioner used to accelerate the solution of \eqref{eq:Ax_equals_b} with an iterative method. The iterative method will typically involve, as a sub-procedure, applying $\mathbf{B}$ to a collection of vectors $\{\mathbf{x}_i\}_{i=1}^m$, thereby generating vectors $\{\mathbf{y}_i = \mathbf{B} \mathbf{x}_i\}_{i=1}^m$. Instead of rebuilding the H-matrix preconditioner for each successive linear system, we recommend \emph{recycling} the information generated by the iterative method while solving the current linear system, in order to update the preconditioner $P$ for the next linear system. In detail, let $\mathbf{X}$ and $\mathbf{Y}$ denote the $N \times m$ matrices with columns given by the vectors $\mathbf{x}_i$ and $\mathbf{y}_i$, respectively, so that the information learned about $\mathbf{B}$ while solving the current linear system may be summarized by the equation:
\begin{equation*}
\mathbf{B}\mathbf{X} = \mathbf{Y}.
\end{equation*}
For convenience, define 
\begin{equation*}
\mathbf{R} := \mathbf{X}-\mathbf{P}\mathbf{Y}, \qquad \text{and} \qquad \boldsymbol{\Theta} := \left(\mathbf{X}^T\mathbf{Y}\right)^{-1}.
\end{equation*}
One of the following three formulas may be used to perform a low rank update of $\mathbf{P}$:
\begin{align}
	\mathbf{P} &\gets \mathbf{P} + \mathbf{R}\left(\mathbf{Y}^T \mathbf{Y}\right)^{-1} \mathbf{Y}^T \label{eq:res_update} \\
	\mathbf{P} &\gets \mathbf{P} + \mathbf{R} \left(\mathbf{R}^T \mathbf{Y}\right)^{-1} \mathbf{R}^T \label{eq:srk_update} \\
	\mathbf{P} &\gets \left(\mathbf{I} - \mathbf{X}\boldsymbol{\Theta}\mathbf{Y}^T\right)\mathbf{P}\left(\mathbf{I} - \mathbf{Y}\boldsymbol{\Theta}\mathbf{X}^T\right) + \mathbf{X}\boldsymbol{\Theta}\mathbf{X}^T \label{eq:dfp_update}
\end{align}
In all three cases, direct calculation shows that we have $\mathbf{P}\mathbf{Y}=\mathbf{X}$ after the update, which implies $\mathbf{P}^{-1} \mathbf{X} = \mathbf{B}\mathbf{X}$. Formulas \eqref{eq:res_update} and \eqref{eq:srk_update} perform rank-$k$ updates of $\mathbf{P}$, while formula \eqref{eq:dfp_update} performs a rank-$2k$ update. Formula \eqref{eq:res_update} does not preserve symmetry, and should therefore be used when $\mathbf{B}$ is not symmetric. Formula \eqref{eq:srk_update} preserves symmetry but not positive definiteness, and should therefore be used when $\mathbf{B}$ is symmetric and indefinite. Formula \eqref{eq:dfp_update} preserves both symmetry and positive definiteness, and should therefore be used when $\mathbf{B}$ is symmetric positive definite. Formula \eqref{eq:srk_update} is known as the ``symmetric rank-$k$'' update formula. Formula \eqref{eq:dfp_update} is known as the DFP update formula, named after it's inventor Davidson, and polularizers Fletcher and Powell. For more details on these update formulas, we recommend the discussion in Chapter 6 of \cite{NocedalWright99}. In \cite{NocedalWright99}, one-vector versions of formulas \eqref{eq:srk_update} and \eqref{eq:dfp_update} are discussed thoroughly; the versions here are straightforward generalizations to many-vector updates.

%to update $\mathbf{B}^{(k)} \approx \left(\mathbf{A}^{(k)}\right)^{-1}$ denote the H-matrix preconditioner used for the $k$th linear system, and set $\mathbf{B}^{(1)}:= \mathbf{A}_H^{-1}$ denote the H-matrix approximation used to precondition the first linear system. Rather than re-build the H-matrix preconditioner for each subsequent linear system, one may instead form a preconditioner, $\mathbf{B}^{(k+1)}$ update, by \emph{recycling} the information generated by the iterative method used to solve the current system.
%
%In practice it is often necessary to solve a sequence of linear systems of the form 
%\begin{equation*}
%	\mathbf{A} \mathbf{x} = \mathbf{b}
%\end{equation*}
%where either $\mathbf{A}$, or $\mathbf{b}$, or both, are slowly varying. 
%
%\begin{equation*}
%	\mathbf{B}\mathbf{X} = \mathbf{Y}
%\end{equation*}
%
%
%
%\begin{equation*}
%\mathbf{Q} := \mathbf{X}\boldsymbol{\Theta}\mathbf{Y}^T
%\end{equation*}
%
%\begin{equation}
%	\mathbf{B}_{k+1} = \left(\mathbf{I} - \mathbf{Y}\boldsymbol{\Theta}\mathbf{X}^T\right)\mathbf{B}_k\left(\mathbf{I} - \mathbf{X}\boldsymbol{\Theta}\mathbf{Y}^T\right) + \mathbf{Y}\boldsymbol{\Theta}\mathbf{Y}^T
%\end{equation}
